---
title: "Converting from and to paleotree format"
author: "Joelle Barido-Sottani"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Converting from and to paleotree format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(FossilSim)
if(!require(paleotree)) knitr::opts_chunk$set(eval = FALSE)
```

## Compatibility with paleotree

Objects produced by `FossilSim` contain all the information required by the `paleotree` format and vice-versa, thus these two packages are fully compatible.

Converting from `paleotree` is done using the `paleotree.record.to.fossils` function, which returns a list of the corresponding `phylo`, `taxonomy` and `fossils` objects.

```{r}
record = paleotree::simFossilRecord(p=0.1, q=0.1,r=0.1, nruns=1, nTotalTaxa=c(30,40),
                                    nExtant=0, nSamp = c(5,25))

converted_record = paleotree.record.to.fossils(record)
converted_record
```

If the `paleotree` record is fully extinct, i.e contains no extant samples, the age of the record will be encoded in the field `origin.time` of the tree.

```{r}
converted_record$tree$origin.time
```

Similarly, converting from the `FossilSim` objects to `paleotree` format is done using the `fossils.to.paleotree.record` function, which requires a `fossils` object and either a `phylo` or `taxonomy` object. If no taxonomy is provided, all speciation events in the tree will be assumed to be symmetric (bifurcating).

```{r}
t = ape::rtree(6)
f = sim.fossils.poisson(rate = 2, tree = t)

converted_record = fossils.to.paleotree.record(f, tree = t)
converted_record
```
