---
title: "Exporting to zero-edge format"
author: "Joelle Barido-Sottani"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{FossilSim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(FossilSim)
```

## Zero-edge format

Many phylogenetic inference softwares, such as [BEAST2](http://beast2.org/) and [RevBayes](https://revbayes.github.io), can handle trees containing fossil samples. In the zero-edge format, tip samples are represented normally in the tree, but sampled ancestors, i.e fossil samples which have sampled descendants, are represented as tips at the end of edges with length 0.

This format is included in `FossilSim` as the `SAtree` objects.

## Converting to SAtree

`SAtree` objects can be built from a `fossils` object and the associated tree. Tips (both extinct and extant) will be labelled with the species they were sampled from followed by an index: the oldest sample for a given species will get index 1 and all other samples will be ordered from oldest to youngest.
The `SAtree` format also includes a field `complete` which indicates whether the tree should be considered as including all lineages of the process or only sampled lineages. Note that if the tree is complete, the youngest tip of a given extinct species represents the extinction event for this species and not a fossil sample.

```{r}
t = ape::rtree(6)
f = sim.fossils.poisson(rate = 2, tree = t)

SAt = SAtree.from.fossils(t,f)
print(SAt)
print(SAt$complete)
```

Note that the default plotting function for `SAtree` objects is `plot.asymmetric.tree` which plots a stratigraphic range representation of the tree and fossil samples, while assuming that all speciation events are asymmetric.

```{r}
plot(SAt)
```

To use a different plotting function, not that `SAtree` objects are also `phylo` objects and can still be used with all the plotting compatible with that format.
```{r}
ape::plot.phylo(SAt)
```

## Other useful functions

Other functions are present in `FossilSim` to modify trees. 
`prune.fossils` will remove all intermediate fossil samples and keep only the first and last occurrences of any species.

```{r}
SAt_pruned = prune.fossils(SAt)
ape::plot.phylo(SAt_pruned)
```

`sampled.tree.from.combined` removes all unsampled lineages from the tree, and can optionally apply a sampling probability to extant samples, or keep only a specified list of extant samples.

```{r}
SAt_sampled = sampled.tree.from.combined(SAt)
ape::plot.phylo(SAt_sampled)
```

These two functions can be applied to any object of type `phylo`, but are designed to work on trees which include fossil samples.
